% Methods-only extract for review
\documentclass[11pt]{article}
\usepackage{acl}
\usepackage{newtxtext}
\usepackage{newtxmath}
\let\Bbbk\relax
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{hyperref}

\title{Methods: Probing Causal Structure in LLM Forecasters}

\author{
  Sean Kelley \and Christoph Riedl \\
  Northeastern University
}

\begin{document}
\maketitle

% ============================================================================
\section{Methodology}
\label{sec:method}

\subsection{Overview}

Our pipeline consists of four stages applied to each forecasting question $q$:

\begin{enumerate}
    \item \textbf{Causal Forecast} (\S\ref{sec:dag}): The LLM produces an initial probability $p_0 \in [0.01, 0.99]$ and a causal DAG $G = (V, E)$ with $k$ factor nodes (default $6 \leq k \leq 10$; $12 \leq k \leq 16$ in the network size condition), 1 outcome node, and directed edges annotated with causal mechanisms.
    \item \textbf{Network Analysis} (\S\ref{sec:network}): We compute betweenness centrality and outcome mediation on $G$ and select ${\sim}$21 probe targets spanning 14 probe types across four categories (strengthen, negate, structural challenge, control).
    \item \textbf{Probe Generation} (\S\ref{sec:probes}): The LLM generates a natural-language counterfactual probe for each target, contextualized to the specific question and network element.
    \item \textbf{Probed Forecast} (\S\ref{sec:probed}): For each probe, we present the LLM (in a fresh single-turn conversation) with the original question, initial probability, full causal network, and probe text, and elicit an updated probability $p_i$.
\end{enumerate}

The key dependent variable is the magnitude of probability revision for each probe $i$. Because the model generates its own DAG in Stage~1, we are not testing whether LLMs perform causal reasoning \emph{de novo}---we are testing whether their probability updates are \textbf{internally consistent} with their own elicited causal structure. A model that places a factor on many paths to the outcome has implicitly committed to that factor's importance; we test whether it honors that commitment when the factor is challenged.

\subsection{Questions}

We draw questions from ForecastBench \cite{zou2024forecastbench}, a benchmark of 500 binary forecasting questions sourced from prediction markets (Metaculus, Polymarket, Manifold, INFER) and data providers (FRED, ACLED, Yahoo Finance, DBnomics). All questions have resolution dates in late 2025, after the training data cutoffs of all evaluated models.

\paragraph{Causal complexity filtering.} Because our pipeline elicits causal DAGs with 6--16 factor nodes, we require questions whose outcomes are shaped by a genuine network of interacting causal factors across multiple domains. We classify all 500 ForecastBench questions using GPT-5.4 (\texttt{openai/gpt-5.4}) with a structured prompt (Appendix~\ref{app:complexity-prompt}) that applies five criteria: (1)~a domain expert would naturally identify $12+$ distinct, non-redundant causal factors that \emph{interact} with each other; (2)~these factors span $3+$ domains (political, economic, social, technological, military, regulatory, environmental, organizational); (3)~the factors form a genuine causal \emph{network} with indirect pathways and mediating variables, not a linear chain; (4)~the question is \emph{grounded} in real, identifiable actors or measurable conditions with current observable evidence; and (5)~the expert reasoning naturally involves understanding how factors interact to shape the outcome, not just checking a sequence of conditions. Questions are classified as LOW if they are primarily answerable via time-series extrapolation, tournament modeling, pipeline reasoning, or market pricing --- though these methodology shortcuts serve as guides rather than hard rules.

The key distinction is whether the \emph{drivers} of the outcome form an interacting multi-domain network. Questions that resolve as a single discrete event (e.g., ``Will country X invade country Y?'') are classified as HIGH if the conditions that determine that event emerge from a complex interacting network of military, diplomatic, economic, and domestic factors. 
This yields 116 high-complexity questions (23.2\% of the pool). These span conflict \& security (40), technology (27), politics \& governance (18), health \& science (13), finance \& markets (6), economics (5), society \& culture (4), and climate \& energy (3). The 384 excluded questions are predominantly time-series extrapolations (stock prices, weather comparisons, financial indicators), single-domain outcomes (sports, entertainment awards), speculative/ungrounded scenarios, and narrow institutional decisions. All 116 questions are shared across models to enable paired comparison.

\subsection{Causal Forecast Prompt Design}
\label{sec:superforecasting}

Our Stage 1 prompt is deliberately minimal: the LLM is instructed to act as an ``expert forecaster,'' estimate a probability, and represent its reasoning as a directed causal graph (see Appendix~\ref{app:prompt-forecast} for the full prompt). The prompt specifies structural requirements---6--10 factor nodes, one outcome node, directed edges with causal mechanisms, indirect chains not just direct links---but does not instruct the model on \emph{how} to reason. There are no instructions about base rates, decomposition strategies, counterfactual testing, or calibration heuristics. This design ensures that the elicited causal structure reflects the model's own reasoning rather than experimenter-imposed scaffolding.

\subsection{Causal DAG Elicitation}
\label{sec:dag}

In Stage 1, we prompt the LLM to produce:
\begin{itemize}
    \item A probability estimate $p_0$
    \item A set of \textbf{factor nodes} $V_F = \{v_1, \ldots, v_k\}$ ($6 \leq k \leq 10$), each with an identifier and natural-language description
    \item An \textbf{outcome node} $v_o$ representing the forecasted event
    \item A set of \textbf{directed edges} $E \subseteq V \times V$, each annotated with a causal mechanism
    \item \textbf{Reasoning} explaining the probability estimate
\end{itemize}

The response is returned as structured JSON. We enforce validation constraints: the graph must contain exactly one outcome node, every factor must connect to at least one other node, the outcome must have at least one incoming edge, and all edge endpoints must reference valid nodes.

\subsection{Network Analysis and Probe Target Selection}
\label{sec:network}

We compute graph-theoretic metrics on the elicited DAG $G$ to identify elements of varying structural importance.

\paragraph{Node importance.} We use \textbf{betweenness centrality} as the primary importance metric:
\begin{equation}
    C_B(v) = \sum_{s \neq v \neq t} \frac{\sigma_{st}(v)}{\sigma_{st}}
\end{equation}
where $\sigma_{st}$ is the number of shortest paths from $s$ to $t$ and $\sigma_{st}(v)$ is the number passing through $v$. Betweenness captures a node's role as a mediator in the causal network---high-betweenness nodes sit on many paths between factors and the outcome. We additionally compute \textbf{outcome mediation}---the fraction of shortest paths from any factor node to the outcome that pass through $v$---which captures a node's role as a mediator specifically on outcome-directed paths. Unlike betweenness (which counts paths between all pairs), outcome mediation is outcome-directed and serves as a continuous measure of how central the node is to the forecasting target.

\paragraph{Edge classification.} For probe target selection, edges are classified by \textbf{shortest-path membership} (whether the edge lies on any shortest path from a factor to the outcome). Shortest-path edges receive both negate and strengthen probes; non-shortest-path (peripheral) edges serve as edge-level baselines.

\paragraph{Probe target selection.} We select ${\sim}$20 targets per question using a structured allocation that spans the importance spectrum (Table~\ref{tab:probe_types}). Factor nodes are sorted by betweenness centrality and each importance tier receives both negate and strengthen probes: the top 2 nodes (high), the median node (medium), and the lowest-centrality node (low). Edges are similarly partitioned: shortest-path edges receive both negate and strengthen probes, as does a non-shortest-path (peripheral) edge. Two control probe types---\texttt{missing\_node} and \texttt{irrelevant}---are included as structural validity checks. High-importance nodes receive two probes each because the top-centrality nodes typically mediate distinct causal paths, providing broader coverage of the network's core structure, while the single lowest-centrality node serves as a peripheral baseline.

\begin{table}[t]
\centering
\small
\begin{tabular}{llp{4.2cm}c}
\toprule
\textbf{Category} & \textbf{Probe Type} & \textbf{Description} & \textbf{$n$} \\
\midrule
\multirow{5}{*}{Strengthen}
  & node\_strengthen (high) & Reinforce top-2 importance nodes & 2 \\
  & node\_strengthen (medium) & Reinforce median node & 1 \\
  & node\_strengthen (low) & Reinforce lowest node & 1 \\
  & edge\_strengthen (critical) & Reinforce edges on shortest path & 2 \\
  & edge\_strengthen (peripheral) & Reinforce edge not on shortest path & 1 \\
\midrule
\multirow{5}{*}{Negate}
  & node\_negate (high) & Challenge top-2 importance nodes & 2 \\
  & node\_negate (medium) & Challenge median node & 1 \\
  & node\_negate (low) & Challenge lowest node & 1 \\
  & edge\_negate (critical) & Challenge edges on shortest path & 2 \\
  & edge\_negate (peripheral) & Challenge edge not on shortest path & 1 \\
\midrule
\multirow{3}{*}{\shortstack[l]{Structural\\Challenge}}
  & edge\_reverse & Argue causal direction is reversed & 1 \\
  & edge\_structural & Propose a missing causal link & 1--2 \\
  & missing\_node & Suggest an omitted factor & 2 \\
\midrule
Control
  & irrelevant & Introduce irrelevant info & 2 \\
\bottomrule
\end{tabular}
\caption{Probe types and allocation per question. Each question receives ${\sim}$21 probes (19--21 depending on graph structure) spanning four intervention categories. The symmetric negate/strengthen design at each importance level enables comparison of disconfirmatory vs.\ confirmatory sensitivity at both node and edge levels. Structural challenge probes test graph architecture rather than variable states. Nodes are ranked by betweenness centrality; edges by shortest-path membership.}
\label{tab:probe_types}
\end{table}

\subsection{Probe Generation}
\label{sec:probes}

For each target, we prompt the LLM with the original question, initial probability, the full causal network, and type-specific instructions (Table~\ref{tab:probe_instructions}). For example, a node negate probe instructs the model to generate a plausible challenge arguing that the target factor is incorrect or irrelevant; a node strengthen probe instructs it to generate evidence reinforcing that factor. Importantly, the probe generation instructions do not reveal the structural importance of the target---the same instruction template is used for high-, medium-, and low-importance nodes, ensuring that any linguistic differences between probes reflect the semantic content of the target rather than experimenter-imposed framing. Structural challenge probes (\texttt{edge\_reverse}, \texttt{edge\_structural}, \texttt{missing\_node}) question the graph architecture rather than variable states. Control probes (\texttt{irrelevant}) are generated freely without a specific structural target.


\subsection{Probed Forecast}
\label{sec:probed}

Each probe is presented in an independent, single-turn conversation. The LLM receives the original question, its initial probability $p_0$, the full causal network (nodes, edges, and mechanisms), and the probe text. It produces an updated probability $p_i \in [0.01, 0.99]$ and reasoning explaining the update. This one-turn design ensures that probe results are independent---each probe is evaluated without knowledge of other probes or their effects.

All stages use temperature 0.7.

\subsection{Statistical Framework}
\label{sec:metrics}

Our primary analysis uses \textbf{linear mixed-effects (LME) regression} (fitted via \texttt{lme4::lmer} in R with Satterthwaite $p$-values from \texttt{lmerTest}) to test whether topological properties of the probed element predict the magnitude of probability revision. This approach uses the full variance of continuous topological predictors while accounting for question-level clustering and model-specific baselines.

\paragraph{Dependent variable.}
Because raw probability shifts are mechanically bounded (a forecast of $p_0 = 0.90$ can shift at most $+0.09$ upward but $-0.89$ downward), we use the \textbf{absolute log-odds shift} as our primary dependent variable:
\begin{equation}
    |\Delta\text{logit}|_i = \left| \log\frac{p_i}{1-p_i} - \log\frac{p_0}{1-p_0} \right|
\end{equation}
This transformation removes boundary compression and places shifts on an unbounded scale. We additionally include \textbf{probe direction} (negate vs.\ strengthen) as a fixed-effect covariate to control for asymmetric updating---models may systematically shift more in response to confirmatory than disconfirmatory evidence.

\paragraph{Model specification.}
We fit models of the general form:
\begin{equation}
    |\Delta\text{logit}|_{mqi} = \beta_0 + \beta_1 \cdot x_{qi}^{(z)} + \beta_2 \cdot \text{Direction}_i + \gamma_m + u_q + \varepsilon_{mqi}
\end{equation}
where $x_{qi}^{(z)}$ is the topological predictor $z$-scored within each question $q$ (to normalize for graph size), $\text{Direction}_i$ is a binary indicator for strengthen vs.\ negate probes, $\gamma_m$ is a model fixed effect, $u_q \sim \mathcal{N}(0, \sigma_q^2)$ is a random intercept by question, and $\varepsilon_{mqi}$ is residual error. The key parameter is $\beta_1$: a positive $\beta_1$ indicates that more structurally important elements elicit larger probability revisions after controlling for probe direction and model-level baselines.

We fit two variants for each predictor: (1) a \textbf{shared-slope} model (single $\beta_1$ across all models, used for headline results) and (2) an \textbf{interaction model} ($\beta_1 \times \text{model}$, yielding model-specific slopes). For each model, we compute the total slope ($\beta_1 + \beta_{1 \times m}$) and its standard error from the variance--covariance matrix, testing against zero to determine whether that model individually exhibits the topological effect.

\paragraph{Data exclusions.}
Questions where all probed nodes have identical betweenness centrality (zero variance after $z$-scoring) are excluded from LME fitting, as the predictor carries no information. This affects 16/116 questions in the edge permutation condition (where random rewiring can produce uniform centrality) and $\leq$2 questions in the main analysis. All DAGs are validated at generation time to ensure no disconnected nodes: any factor with zero edges is rejected and regenerated.

\paragraph{Topological predictors.}
We test two node-level predictors, each in a separate LME model:

\begin{itemize}
    \item \textbf{Node betweenness centrality} ($C_B$): fraction of all shortest paths passing through the node. This is a generic graph centrality measure---high-betweenness nodes mediate many paths between pairs of nodes in the network.
    \item \textbf{Outcome mediation}: fraction of factor$\rightarrow$outcome shortest paths passing through the node. Unlike betweenness (which counts paths between all pairs), outcome mediation is outcome-directed and captures how central the node is specifically to the forecasting target.
\end{itemize}

\paragraph{Probe taxonomy.}
Probes are classified into four intervention categories: \textbf{Strengthen} (reinforce existing causal claims), \textbf{Negate} (challenge existing claims), \textbf{Structural Challenge} (challenge the graph architecture---e.g., propose missing edges, reverse edge direction, suggest omitted nodes), and \textbf{Control} (irrelevant information). Within each category, probes target either node-level or edge-level elements, enabling factorial comparison.

\paragraph{Coherence checks.}
We assess reasoning coherence via three supplementary analyses: (1) \textbf{stated-impact ratings}: a GPT-4o-mini judge rates each probe response's stated impact on a 1--5 scale, which we correlate with actual $|\Delta|$; (2) \textbf{Bayesian coherence}: log-odds shift as a function of initial probability; (3) \textbf{embedding separation}: within-question cosine similarity of reasoning embeddings (\texttt{text-embedding-3-large}) for targeted probes vs.\ irrelevant probes.

\paragraph{Cross-model DAG agreement.}
Because models assign different names to semantically equivalent factors (e.g., \texttt{economic\_growth} vs.\ \texttt{economic\_conditions}), exact string matching underestimates structural agreement. We therefore align nodes across models using embedding-based semantic matching before computing graph edit distance.

For each question and model pair, we embed every node as ``\texttt{id}: \texttt{description}'' using \texttt{text-embedding-3-large}, construct a cosine-similarity matrix, and find the optimal one-to-one node alignment via the Hungarian algorithm \citep{kuhn1955hungarian}, counting pairs with cosine similarity $\geq 0.7$ as matched. We then compute \emph{semantic normalized graph edit distance} (nGED): unmatched nodes from each graph count as node edit operations, and edges are compared structurally by remapping one graph's edges through the node alignment and taking the symmetric difference. Formally, $\text{nGED} = (\text{node\_ops} + \text{edge\_ops}) \,/\, (|V_1 \cup V_2| + |E_1 \cup E_2'|)$, where $E_2'$ denotes the remapped edge set. Values range from 0 (identical) to 1 (disjoint).

We test via a permutation test ($n = 2{,}000$): for each permutation, we independently shuffle question assignments within each model and recompute pairwise nGED on the mismatched pairs. Cross-model DAGs are significantly more similar for the same question than expected by chance (nGED $= 0.83$, null $= 1.01 \pm 0.001$, $p < .001$), confirming that models converge on similar causal representations for a given forecasting problem.

% ============================================================================
\section{Experimental Setup}
\label{sec:setup}

\subsection{Models}

We evaluate seven models spanning five architecture families and a range of scales, all accessed via OpenRouter:
\begin{itemize}
    \item \textbf{Llama 3.1 8B Instruct} (8B parameters; released Jul 2024; training cutoff Dec 2023)
    \item \textbf{Llama 3.3 70B Instruct} (70B; released Dec 2024; training cutoff Dec 2023)
    \item \textbf{DeepSeek V3} (685B mixture-of-experts, 37B active; released Mar 2025)
    \item \textbf{Qwen3 235B} (235B MoE, 22B active; released Jul 2025)
    \item \textbf{Qwen3 32B} (32.8B dense; released Apr 2025; training cutoff Mar 2025)
    \item \textbf{Gemini 2.5 Flash Lite} (released Jul 2025; training cutoff Jan 2025)
    \item \textbf{GPT-OSS 120B} (117B MoE; released Aug 2025; training cutoff Jun 2024)
\end{itemize}
Training data cutoffs are reported where officially documented; DeepSeek does not publish a cutoff date. All ForecastBench questions have resolution dates in late 2025, after all known cutoffs. Supplementary analyses (edge permutation, structural ablation, network size, temperature sensitivity, spurious context, propagation decay, and superforecasting comparison) use GPT-OSS 120B as the testbed model.

All models use temperature 0.7 for all stages (causal forecast, probe generation, and probed forecast).

\subsection{Data Collection}

We run the full four-stage pipeline on all 116 questions for each of the seven models, yielding approximately ${\sim}$2,400 probes per model (${\sim}$17,000 total probe responses). Each question--model pair produces an independent causal DAG, set of probes, and probe results.

\subsection{Interactive Explorer}

We release an interactive web application, the \textbf{Causal Forecast Lab}, that allows users to browse the elicited DAGs, probe texts, and probability shifts for every question--model combination. The explorer visualizes the causal network with D3.js force-directed layouts, highlights probe targets by importance tier, and displays shift magnitudes alongside model reasoning. This tool supports qualitative inspection of individual cases and complements the aggregate statistical analyses reported here.

\subsection{Robustness Analyses}

We conduct the following supplementary analyses, all using GPT-OSS 120B as the testbed model on the same 116 high-complexity questions unless otherwise noted:

\paragraph{Test--retest reliability.} We run Stage~1 (causal forecast: probability + DAG) twice on the same 116 questions for all seven models and correlate the 116 paired initial probabilities across runs using Spearman $\rho$.

Initial probability estimates are highly reproducible across all models: Spearman $\rho$ ranges from 0.725 (Llama~8B) to 0.960 (Qwen3~235B), with mean absolute error between 0.025 and 0.094. No model shows systematic drift in mean probability across runs (all within $\pm$0.03). The full results are: Qwen3~235B ($\rho = 0.960$), Llama~70B ($\rho = 0.944$), Gemini~Flash~Lite ($\rho = 0.872$), DeepSeek~V3 ($\rho = 0.847$), GPT-OSS ($\rho = 0.791$), Qwen3~32B ($\rho = 0.789$), and Llama~8B ($\rho = 0.725$).

\paragraph{Temperature sensitivity.} We run Stage 1 (causal forecast) at temperatures $T \in \{0.0, 0.3, 0.7, 1.0\}$ on 116 questions using GPT-OSS 120B, with $T = 0.7$ as the pipeline default. For each temperature we compute the mean initial probability with a 95\% bootstrap confidence interval (10,000 resamples). For each pair of temperatures we compute the \textbf{normalized graph edit distance} (nGED) between same-question DAGs, averaged across all 116 questions:
\begin{equation}
    \text{nGED}(G_1, G_2) = \frac{|V_1 \triangle V_2| + |E_1 \triangle E_2|}{|V_1 \cup V_2| + |E_1 \cup E_2|}
\end{equation}
where $\triangle$ denotes symmetric difference. nGED ranges from 0 (identical graphs) to 1 (completely disjoint). This measures the fraction of graph elements (nodes and edges) that must be inserted or deleted to transform one DAG into the other.

\paragraph{Edge permutation condition.} For each of the 116 questions, we construct a permuted DAG by preserving the original nodes (same factor names and descriptions) but randomly rewiring the edges (maintaining graph density and DAG constraints). We then present this permuted DAG to GPT-OSS 120B as if it were the model's own reasoning and re-run the full probing pipeline. We fit the same shared-slope LME models (betweenness $\rightarrow |\Delta|$ and outcome mediation $\rightarrow |\Delta|$) on the permuted data and compare $\beta_1$ to the originals. Because the node descriptions are unchanged, the probe text still challenges real causal factors---any residual effect reflects content-driven importance (some factors are inherently more consequential regardless of graph position). A reduction in $\beta_1$ isolates the structure-driven component: the degree to which the model's sensitivity tracks the \emph{topological role} of each factor, not just its semantic content.

\paragraph{Structural ablation.} For each of the 116 questions, we present the model with its full DAG and instruct it to disregard a specific factor node when re-estimating the probability. Unlike physical node removal (which allows the model to silently compensate by re-reasoning from scratch), this ``ignore this factor'' framing keeps the full network visible and tests whether the model can differentially weight factors within its own causal structure. We ablate each factor node independently (node-only; edge ablation is omitted as edges lack a natural importance metric after dropping edge betweenness). We fit LME models predicting $|\Delta|$ from node betweenness and outcome mediation, with random intercepts by question.

\paragraph{Spurious context control.} We test whether the DAG elicitation process is robust to irrelevant background information. For each of the 116 questions, we prepend a factual but topically unrelated paragraph (e.g., medieval Scandinavian runestones for a stock price question, deep-sea hydrothermal vent biology for a geopolitics question). Backgrounds are generated by GPT-4o-mini and independently verified as irrelevant by a second GPT-4o-mini call. We then run Stage 1 on the augmented prompts and use a GPT-4o-mini judge to evaluate whether any DAG node is semantically derived from the spurious background. We additionally compare the resulting probabilities and node sets to the original (unaugmented) DAGs.

\paragraph{Network size.} We vary the Stage 1 prompt across two conditions: 6--10 factor nodes (the default) and 12--16 nodes. Both conditions use the same 116 high-complexity questions. The causal complexity filter (see \S\ref{sec:method}) ensures that all questions can naturally sustain 12+ distinct causal factors---without this filter, narrow-domain questions (e.g., weekly temperature comparisons) fail to produce valid large DAGs because the model cannot generate 12+ meaningful, non-redundant factors. We additionally increase the maximum response length from 2,000 to 4,000 tokens for the 12--16 condition, as valid 14--16 node DAGs with edge descriptions require ${\sim}$1,900 tokens on average. For each condition, we fit the shared-slope LME models (betweenness $\rightarrow |\Delta|$ and outcome mediation $\rightarrow |\Delta|$) and compare the grounding coefficients $\beta_1$ across graph sizes. Larger networks provide more intermediate nodes with varying betweenness, potentially enabling finer-grained importance discrimination.

\paragraph{Factor ranking.} To test whether the structural sensitivity revealed by probing reflects explicit or implicit knowledge, we present each model with its own Stage~1 factor nodes (IDs and descriptions) and ask it to rank them by importance to the outcome. Because the model ranks the exact nodes it generated, no embedding-based matching is required. We compute Spearman $\rho$ between stated rank and betweenness centrality across all factor--rank pairs. A significant negative correlation indicates that the model's explicit importance judgments align with DAG topology (convergent validity); a null result would suggest structural sensitivity is implicit (revealed through probing behavior but not accessible to introspection).

\paragraph{Persuasiveness control.} Because probe generation targets nodes of varying structural importance, probes challenging central nodes may be inherently more persuasive than those challenging peripheral nodes---not due to experimenter framing (which is held constant; see \S\ref{sec:probes}) but because central topics admit stronger arguments. To rule out persuasiveness as a confound, an independent GPT-4o-mini judge---blind to node importance---rated the persuasiveness of 400 probe texts (200 high-importance, 200 low-importance) on a 1--5 scale (Appendix~\ref{app:prompt-persuasiveness}). High- and low-importance probes did not differ significantly in persuasiveness (high: $M = 3.62$, low: $M = 3.53$; Mann--Whitney $U = 21{,}365$, $p = .17$), ruling out probe text quality as a confound for the importance--shift relationship.

\paragraph{Propagation decay.} Using GPT-OSS 120B, we test whether the model's causal reasoning respects graph locality by measuring how probe effects attenuate with graph distance. For each of the 116 questions, we take the existing node-level probes (\texttt{negate\_high}, \texttt{negate\_low}, \texttt{strengthen}, \texttt{strengthen\_low}) and re-prompt the model with the full DAG and probe text, asking it to rate how each \emph{other} factor node's causal influence would change on a $[-1, +1]$ scale ($-1$ = strongly weakened, $+1$ = strongly reinforced, $0$ = unaffected). We annotate each rated node with its undirected shortest-path distance from the probed node and report (i) mean $|\text{impact}|$ and (ii) fraction of affected nodes (impact $\neq 0$) as a function of graph distance. We test for monotonic decay using Spearman $\rho$ between distance and $|\text{impact}|$ across all node--probe pairs.

\paragraph{Superforecasting prompt comparison.} We run Stage 1 with two system prompts on the same 116 questions: (1) the baseline prompt (used in the main pipeline) and (2) a superforecasting-augmented prompt that explicitly instructs outside-view thinking, granular decomposition, signal--noise discrimination, counterfactual validation, and calibration. We compare the resulting DAGs (node/edge Jaccard similarity, graph density, number of nodes) and probability estimates (Spearman $\rho$, mean absolute difference) to assess whether superforecasting instructions meaningfully alter the elicited causal structure.

% ============================================================================
\appendix

\section{System Prompts}
\label{app:prompts}

We provide the full system prompts used at each stage of the pipeline.

\subsection{Stage 1: Causal Forecast}
\label{app:prompt-forecast}

The system prompt instructs the LLM to act as an expert forecaster and produce a probability estimate alongside a causal DAG:

\begin{quote}
\small
\texttt{You are an expert forecaster. Your task is to estimate the probability that a given event will occur, and to represent your reasoning as a directed causal graph.}

\texttt{Build a causal network where:}
\begin{itemize}
\small
\item \texttt{Nodes are factors that influence the outcome}
\item \texttt{Directed edges represent causal mechanisms (A -> B means A causally influences B)}
\item \texttt{One node is the ``outcome'' node representing the forecasted event}
\item \texttt{All other nodes are ``factor'' nodes representing causal drivers}
\end{itemize}

\texttt{You must be well-calibrated: assign probabilities that genuinely reflect your uncertainty. Avoid defaulting to 50\% unless you truly have no information.}

\texttt{Respond with ONLY valid JSON. No other text.}
\end{quote}

The user prompt specifies the question, required JSON schema (probability, nodes with IDs/descriptions/roles, edges with from/to/mechanism, reasoning), and structural constraints:
\begin{itemize}
\small
\item 6--10 factor nodes plus exactly 1 outcome node
\item Each factor must have at least 1 outgoing edge
\item The outcome must have at least 1 incoming edge
\item All edge endpoints must reference valid node IDs
\item Indirect causal chains are required (not just direct links to outcome)
\item Node IDs use \texttt{short\_snake\_case} format
\end{itemize}

\subsection{Stage 2: Probe Generation}
\label{app:prompt-probe-gen}

The system prompt for probe generation:

\begin{quote}
\small
\texttt{You are an expert at constructing probes to test how a forecaster updates their beliefs when elements of their causal reasoning are challenged. You will be given a causal network (nodes and edges) and asked to create a targeted probe that challenges a specific structural element of that network.}

\texttt{The probe should be plausible, specific, and designed to test whether the forecaster can appropriately update their beliefs when confronted with new information about their causal model.}

\texttt{Respond with ONLY valid JSON. No other text.}
\end{quote}

The user prompt provides the question, initial probability, full causal network, and type-specific instructions. Table~\ref{tab:probe_instructions} summarizes the instructions for each probe type.

\begin{table*}[t]
\centering
\small
\begin{tabular}{lp{12cm}}
\toprule
\textbf{Probe Type} & \textbf{Instruction Summary} \\
\midrule
node\_negate (high) & ``Create a direct negation of the causal factor [X]. Provide a specific, plausible argument for why this factor might not matter or why the opposite might be true.'' \\
node\_negate (medium) & ``Create a direct negation of the causal factor [X]. Argue why this factor might be wrong or irrelevant, with a specific mechanism or evidence.'' \\
node\_negate (low) & ``Create a direct negation of the causal factor [X]. Argue why this factor might be wrong or irrelevant.'' \\
node\_strengthen (high) & ``Create evidence that REINFORCES the causal factor [X], making it more likely to be true and more causally important. Provide a strong new data point or development that supports this factor's role in the causal network.'' \\
node\_strengthen (medium) & ``Create evidence that REINFORCES the causal factor [X], making it more likely to be true. Provide a specific piece of evidence or development that supports this factor.'' \\
node\_strengthen (low) & ``Create evidence that REINFORCES the causal factor [X], making it more likely to be true. Provide evidence that supports this factor.'' \\
edge\_negate (critical) & ``Challenge the causal link [X]. Argue why this causal mechanism might be broken, spurious, or much weaker than assumed. Provide a specific counter-mechanism or evidence.'' \\
edge\_negate (peripheral) & ``Challenge the causal link [X]. Argue why this causal mechanism might not hold.'' \\
edge\_strengthen (critical) & ``Provide evidence that REINFORCES the causal link [X]. Argue why this mechanism is even stronger than assumed, with a specific piece of evidence or recent development.'' \\
edge\_strengthen (peripheral) & ``Provide evidence that REINFORCES the causal link [X]. Argue why this mechanism is stronger than assumed.'' \\
\midrule
\multicolumn{2}{l}{\textit{Structural challenge probes}} \\
edge\_reverse & ``Argue that the causal link [X] actually runs in the OPPOSITE DIRECTION. Provide a specific mechanism for reverse causation.'' \\
edge\_structural & ``The network does NOT include a direct link [X]. Argue that this missing link DOES exist and is important. Describe a specific causal mechanism.'' \\
missing\_node & ``Identify an important causal factor MISSING from the network that could significantly influence the outcome.'' \\
\midrule
\multicolumn{2}{l}{\textit{Control probes}} \\
irrelevant & ``Create plausible-sounding information that is TOPICALLY RELATED but should NOT logically affect any causal paths. It should sound relevant at first but have no causal bearing.'' \\
\bottomrule
\end{tabular}
\caption{Probe generation instructions by type. Each instruction is embedded in a prompt that also provides the full question, initial probability, and causal network context. [X] denotes the target node or edge ID and description. Crucially, instructions do not reveal the structural importance of the target to the probe-generating model. While the level of instructional detail varies (e.g., high-importance negate probes request a more specific argument), no instruction mentions the target's centrality, importance rank, or position in the network.}
\label{tab:probe_instructions}
\end{table*}

\subsection{Stage 3: Probed Forecast}
\label{app:prompt-probed}

The system prompt for the probed forecast stage:

\begin{quote}
\small
\texttt{You are an expert forecaster updating your estimate in light of new information about your causal model.}

\texttt{When presented with new information, consider how it affects your causal network and update your probability estimate accordingly.}

\texttt{Respond with ONLY valid JSON. No other text.}
\end{quote}

The user prompt provides the original question, initial probability $p_0$, the full causal network (nodes and edges with mechanisms), and the probe text. The model outputs \texttt{updated\_probability}, \texttt{shift\_direction}, and \texttt{reasoning}. The prompt does not indicate which elements are structurally important or instruct the model to scale updates by importance---any differential sensitivity to structural importance must arise from the model's own assessment of the DAG.

\subsection{Propagation Decay Prompt}
\label{app:prompt-propagation}

The system prompt for the propagation decay analysis:

\begin{quote}
\small
\texttt{You are an expert forecaster reasoning about how new information propagates through a causal network.}

\texttt{When presented with a challenge to one element of your causal network, consider how the effect ripples through to OTHER factors in the network---not just the final outcome. Some factors may be directly downstream of the challenged element; others may be unaffected.}

\texttt{Rate the impact on each factor using a scale from $-$1.0 (strongly weakened) to +1.0 (strongly reinforced), with 0.0 meaning no effect.}

\texttt{Respond with ONLY valid JSON. No other text.}
\end{quote}

The user prompt provides the original question, initial probability, the full causal network, and the probe text from the main pipeline. The model is asked to rate each \emph{other} factor node's impact and provide one-sentence reasoning per node, plus an updated probability. The prompt instructs: ``factors directly downstream of [target] in the network should be affected more than distant or unrelated factors.''

\section{Judge Prompts}
\label{app:judge-prompts}

We use GPT-4o-mini as an independent judge for three validation tasks. All judges are blind to importance labels.

\subsection{Persuasiveness Judge}
\label{app:prompt-persuasiveness}

\begin{quote}
\small
\texttt{You are evaluating the persuasiveness of arguments. Rate how persuasive the following text is on a scale of 1-5:}

\texttt{1 = Very weak, vague, or unconvincing}\\
\texttt{2 = Somewhat weak, lacks specificity}\\
\texttt{3 = Moderately persuasive}\\
\texttt{4 = Strong, specific, and well-argued}\\
\texttt{5 = Very compelling, with concrete evidence or strong reasoning}

\texttt{Respond with ONLY a single integer (1-5). No other text.}
\end{quote}

The user prompt provides the probe text without any importance or probe-type metadata.

\subsection{Stated-Impact Judge}
\label{app:prompt-reasoning-judge}

\begin{quote}
\small
\texttt{You are evaluating a forecaster's reasoning about how a challenge to their causal model affects their probability estimate.}

\texttt{Rate how impactful the forecaster SAYS the challenge is, on a scale of 1-5:}

\texttt{1 = Says the challenge is irrelevant, peripheral, or has no bearing on the outcome}\\
\texttt{2 = Says the challenge is minor, has limited effect, or only weakly affects the estimate}\\
\texttt{3 = Says the challenge is moderate, affects some causal paths but not the main ones}\\
\texttt{4 = Says the challenge is significant, disrupts important causal paths}\\
\texttt{5 = Says the challenge is fundamental, breaks core reasoning, or invalidates key assumptions}

\texttt{Focus ONLY on what the forecaster STATES about the challenge's impact, not on the actual probability shift. Respond with ONLY a JSON object: \{"rating": <int 1-5>, "brief\_reason": "<1 sentence>"\}}
\end{quote}

The user prompt provides the probe text and the model's reasoning response.

\subsection{Uncertainty Judge}
\label{app:prompt-uncertainty-judge}

\begin{quote}
\small
\texttt{You are evaluating the level of epistemic uncertainty expressed in a forecaster's reasoning about a probability estimate.}

\texttt{Rate how much uncertainty the forecaster EXPRESSES in their reasoning, on a scale of 1-5:}

\texttt{1 = Very confident: uses definitive language, no hedging, presents conclusions as certain}\\
\texttt{2 = Mostly confident: occasional qualifiers but overall assertive tone}\\
\texttt{3 = Balanced: acknowledges uncertainty while still taking a position}\\
\texttt{4 = Uncertain: frequent hedging, multiple caveats, expresses doubt about conclusions}\\
\texttt{5 = Very uncertain: dominated by uncertainty language, unable to commit to a position}

\texttt{Focus ONLY on the VERBAL expression of uncertainty, not on the probability value itself. Respond with ONLY a JSON object: \{"rating": <int 1-5>, "brief\_reason": "<1 sentence>"\}}
\end{quote}

\section{Probe Type Examples}
\label{app:probe-examples}

Table~\ref{tab:probe_examples} illustrates all 14 probe types applied to a single question: \emph{``Will Cynthia Erivo win Best Actress at the 98th Academy Awards?''} (Llama-3.3-70B, initial $P(\text{Yes}) = 0.42$). Each row shows one probe with a summary of its text, the updated probability, and the resulting shift.

\begin{table*}[t]
\centering
\small
\begin{tabular}{llccc}
\toprule
\textbf{Probe Type} & \textbf{Example Probe (abbreviated)} & \textbf{$p_0$} & \textbf{$p_1$} & \textbf{Shift} \\
\midrule
\multicolumn{5}{l}{\textit{Node probes}} \\
Negate (high imp.) & Pre-award hype often leads to voter fatigue\ldots & 0.42 & 0.28 & $-$0.14 \\
Negate (med imp.) & Several strong Best Actress contenders have withdrawn\ldots & 0.42 & 0.58 & +0.16 \\
Negate (low imp.) & Academy voters often prioritize newcomers\ldots & 0.42 & 0.38 & $-$0.04 \\
Strengthen (high imp.) & Renowned critic publishes glowing review of Erivo\ldots & 0.42 & 0.58 & +0.16 \\
Strengthen (med imp.) & Emma Stone will also be competing for the award\ldots & 0.42 & 0.31 & $-$0.11 \\
Strengthen (low imp.) & Erivo nominated for several prestigious pre-Oscar awards\ldots & 0.42 & 0.58 & +0.16 \\
\midrule
\multicolumn{5}{l}{\textit{Edge probes}} \\
Negate (shortest-path) & Academy often prioritizes performances that are not popular\ldots & 0.42 & 0.31 & $-$0.11 \\
Negate (peripheral) & Academy voters decide based on performances, not reviews\ldots & 0.42 & 0.38 & $-$0.04 \\
Strengthen (shortest-path) & Erivo's stage performance elevates her Oscar narrative\ldots & 0.42 & 0.53 & +0.11 \\
Strengthen (peripheral) & Academy voters heavily influenced by critical consensus\ldots & 0.42 & 0.55 & +0.13 \\
Reverse & Award buzz often precedes film release, not the reverse\ldots & 0.42 & 0.38 & $-$0.04 \\
Structural (missing edge) & Popular acclaimed films failed to secure nominations\ldots & 0.42 & 0.38 & $-$0.04 \\
\midrule
\multicolumn{5}{l}{\textit{Structural and control probes}} \\
Missing node & Erivo's performance compared to a previous iconic portrayal\ldots & 0.42 & 0.31 & $-$0.11 \\
Control (irrelevant) & The ceremony will feature a new streaming platform\ldots & 0.42 & 0.42 & +0.00 \\
\bottomrule
\end{tabular}
\caption{All 14 probe types applied to a single exemplar question (Llama-3.3-70B). Node and edge probes are grouped by structural importance. Strengthen probes do not always increase $p_1$: strengthening a competing factor (e.g., competition strength) correctly lowers the target probability.}
\label{tab:probe_examples}
\end{table*}

\section{Probe Target Selection Algorithm}
\label{app:target-selection}

The probe target selection algorithm operates on the output of network analysis and allocates targets as follows:

\begin{enumerate}
\item \textbf{Sort factor nodes} by betweenness centrality (descending), excluding the outcome node.
\item \textbf{Node targets:}
    \begin{itemize}
    \item Top 2 nodes $\rightarrow$ \texttt{negate\_high} (2) + \texttt{strengthen} (2)
    \item Median node $\rightarrow$ \texttt{negate\_medium} (1) + \texttt{strengthen\_medium} (1)
    \item Bottom node $\rightarrow$ \texttt{negate\_low} (1) + \texttt{strengthen\_low} (1)
    \end{itemize}
\item \textbf{Partition edges} into shortest-path (on any shortest path from a factor to the outcome) and peripheral.
\item \textbf{Edge targets:}
    \begin{itemize}
    \item Top 2 shortest-path edges by edge betweenness $\rightarrow$ \texttt{negate\_critical} (2) + \texttt{strengthen\_critical} (2)
    \item Top peripheral edge $\rightarrow$ \texttt{negate\_peripheral} (1) + \texttt{strengthen\_peripheral} (1)
    \item Top shortest-path edge $\rightarrow$ \texttt{reverse} (1)
    \end{itemize}
\item \textbf{Spurious edges:} Select up to 2 non-adjacent node pairs (no existing edge) for \texttt{edge\_structural} probes.
\item \textbf{Control probes:} 2 \texttt{missing\_node} + 2 \texttt{irrelevant} (generated freely by the LLM without a specific target).
\end{enumerate}

The exact count varies by graph structure (19--21 probes, mean ${\sim}$21) because some graphs have fewer shortest-path edges or non-adjacent node pairs.

\section{Metric Definitions and Statistical Tests}
\label{app:metrics}

\paragraph{Bootstrap confidence intervals.} Descriptive metrics (mean shifts by category) are reported with 95\% bootstrap confidence intervals using 10,000 resamples. LME models report Wald confidence intervals and Satterthwaite $p$-values.

\paragraph{Test--retest reliability.} We report Spearman $\rho$ between paired initial probabilities across two independent Stage~1 runs. Per-model results ($n = 116$ questions each): Qwen3-235B ($\rho = .960$), Llama-70B ($\rho = .944$), Gemini-Flash-Lite ($\rho = .872$), DeepSeek-V3 ($\rho = .847$), GPT-OSS ($\rho = .791$), Qwen3-32B ($\rho = .789$), Llama-8B ($\rho = .725$). Mean absolute error ranges from 0.025 to 0.094. For DAG structural stability, we compute semantic nGED between same-model test--retest DAGs: DeepSeek-V3 (0.62), Llama-70B (0.63), Gemini-Flash-Lite (0.65), Qwen3-235B (0.66), GPT-OSS (0.69), Qwen3-32B (0.77), Llama-8B (0.82). Larger models produce more stable causal structures across runs, though all models show substantially more within-model consistency than the cross-model null ($\text{nGED} = 1.01$). Qualitatively, larger models' DAG variation is primarily cosmetic (paraphrased node names, factor splitting/merging), while smaller models construct genuinely different causal decompositions---yet probability estimates remain more stable than DAG structure, suggesting models converge on similar conclusions through varying causal framings.

\paragraph{Temperature sensitivity results.} Mean initial probabilities are stable across temperatures (range 0.33--0.34). Pairwise semantic nGED increases monotonically with temperature distance: adjacent temperatures differ by nGED $\approx 0.67$--$0.70$, while the most distant pair ($T = 0.0$ vs.\ $T = 1.0$) reaches nGED $= 0.71$. All values remain well below the cross-model null ($\text{nGED} = 1.01$), confirming that temperature-induced variation is far smaller than inter-model structural divergence.

\section{Additional Analysis Details}
\label{app:additional}

\subsection{Spurious Context Control}

We test whether prepending irrelevant background information contaminates DAG construction. For each of the 116 questions, we pair the original question with a factual paragraph drawn from an unrelated domain (e.g., the biomechanics of woodpecker skull shock absorption, competitive pigeon racing in Belgium, the archaeology of Nabataean water management at Petra). Topics are drawn from a pool of obscure domains and assigned to questions such that no topical overlap exists. Each background is generated by GPT-4o-mini conditioned on the assigned topic and verified as irrelevant by a separate GPT-4o-mini call; any flagged pair is regenerated with a different topic.

We run Stage 1 (causal forecast) on the augmented prompts using GPT-OSS 120B and evaluate incorporation with a GPT-4o-mini judge that examines each DAG node for semantic relatedness to the spurious background.

\paragraph{Background generation prompt.} Each background is generated with the following GPT-4o-mini system prompt:

\begin{quote}
\small
\texttt{You generate factual background paragraphs about specific topics. The paragraph must: (1) Be 3--4 sentences with specific numerical facts and measurements. (2) Be entirely about the assigned topic. (3) Contain NO connection whatsoever to the forecasting question shown. (4) Sound like an encyclopedia excerpt.}
\end{quote}

\paragraph{Verification prompt.} A separate GPT-4o-mini call verifies irrelevance:

\begin{quote}
\small
\texttt{You are checking whether a background paragraph could be considered relevant to a forecasting question. Be strict: flag ANY possible connection, even indirect or metaphorical. Consider: direct topical overlap, indirect relevance, metaphorical/analogical connections, geographic overlap, temporal overlap.}
\end{quote}

\paragraph{Incorporation judge prompt.} After DAG construction, a GPT-4o-mini judge evaluates contamination:

\begin{quote}
\small
\texttt{You are evaluating whether a causal DAG has been contaminated by irrelevant background information. You will be given: (1) a forecasting question, (2) irrelevant background information that was presented alongside the question, (3) the nodes of a causal DAG the model constructed. Your task: determine whether ANY node in the DAG was influenced by or derived from the irrelevant background information. A node counts as ``incorporated'' if its concept, topic, or reasoning is semantically related to the background---even if the exact words differ.}
\end{quote}

\paragraph{Results.}
Across all 116 questions, \textbf{zero} DAG nodes are derived from the spurious background (0/116 contamination rate). Probability estimates remain highly correlated with those from unaugmented DAGs (Spearman $\rho = 0.728$, MAE $= 0.085$), confirming that models construct topically appropriate causal networks despite the irrelevant preamble.

\subsection{Superforecasting Prompt Comparison}

We compare DAGs produced by two Stage 1 system prompts on the same 116 questions using GPT-OSS 120B:

\paragraph{Baseline prompt.} The standard prompt used in the main pipeline, which instructs the LLM to ``be well-calibrated'' and produce a causal network (Appendix~\ref{app:prompt-forecast}).

\paragraph{Superforecasting-augmented prompt.} Adds five explicit principles before the network construction instructions:
\begin{enumerate}
\small
\item \textit{Outside view first:} ``Consider the base rate. How often do events like this happen historically? Start from reference classes before incorporating case-specific details.''
\item \textit{Granular decomposition:} ``Break the problem into independent sub-components. Each node should represent a distinct, measurable factor.''
\item \textit{Signal--noise discrimination:} ``Only include factors that have genuine causal relevance. Resist the temptation to add factors just because they are topically related.''
\item \textit{Counterfactual validation:} ``For each causal link, ask: `If this factor were different, would the outcome probability actually change?' ''
\item \textit{Calibration:} ``Assign probabilities that genuinely reflect your uncertainty. Use historical base rates as anchors.''
\end{enumerate}

We compare the two conditions on: probability Spearman $\rho$ ($= 0.713$), mean absolute probability difference ($0.096$), semantic nGED ($0.714$, permutation $p < .001$, null $= 0.998$), mean node count (baseline $8.4$, SF $8.3$; Wilcoxon $p = 0.026$), mean edge count ($8.6$ vs $8.1$; $p = 0.008$), and graph density ($0.139$ vs $0.136$; $p = 0.446$). Same-question DAGs are significantly more similar than cross-question chance, but the superforecasting prompt induces more structural change than test--retest variation (nGED $= 0.71$ vs.\ $0.69$ for GPT-OSS retest). Probability estimates remain correlated, suggesting that explicit superforecasting instructions alter the causal decomposition more than the final probability estimate.

\subsection{Factor Ranking: Convergent and Incremental Validity}
\label{app:factor-ranking}

We present each model with its own Stage~1 factor nodes and ask it to rank them by importance to the outcome (see \S\ref{sec:setup} for procedure). Because the model ranks the exact node IDs it generated, no embedding-based matching is required. We fit LME models with random intercepts by question and model fixed effects across all 7 models ($N = 6{,}495$ node--probe observations, 116 questions).

\paragraph{Convergent validity.} Stated rank strongly predicts betweenness centrality ($\beta = -0.336$, $t = -29.26$, $p < .001$, $R^2_m = .154$): models that rank a factor as most important also place it centrally in their causal topology. The negative sign reflects the rank scale (rank 1 = most important, highest betweenness).

\paragraph{Incremental validity.} Both stated rank and betweenness centrality independently predict probe sensitivity ($|\Delta\text{logit}|$). Stated rank alone achieves $R^2_m = .243$ ($\beta = -0.066$, $t = -20.66$); betweenness alone achieves $R^2_m = .219$ ($\beta = 0.045$, $t = 13.64$). In the combined model, betweenness remains significant after controlling for stated rank ($\beta = 0.025$, $t = 7.30$, $p < .001$), increasing $R^2_m$ to $.249$. This indicates that the DAG topology captures implicit structural information about how the model will respond to probes---information not fully accessible through explicit introspective ranking. The graph structure encodes relational properties (path mediation, bottleneck position) that a flat importance ranking cannot represent.

\input{figures/lme_factor_ranking_summary_table}

\subsection{Question Difficulty Rating}
\label{app:difficulty}

A question's initial probability $p_0$ reflects how likely an event is, but not how \emph{hard} it is to forecast accurately.
A 50\% estimate may reflect well-understood equipoise or deep epistemic opacity; a 95\% estimate may be trivially easy or hinge on opaque insider dynamics.
To disentangle forecasting difficulty from the base rate, we construct an independent difficulty ranking using pairwise LLM-as-judge comparison.
Conceptually, this ranking captures a question's position on the aleatoric--epistemic spectrum: easy questions tend to involve well-understood stochastic processes (weather, economic indicators) with primarily aleatoric uncertainty, while hard questions involve novel situations with opaque causal mechanisms (geopolitical contingencies, long-horizon technology forecasts) where epistemic uncertainty dominates.

\paragraph{Tournament design.}
We run a Swiss-system tournament over the 116 questions using GPT-4o-mini as judge.
In each round, questions with similar current ratings are paired; the judge sees both question texts and selects which is harder to predict accurately, considering information availability, causal complexity, temporal sensitivity, domain opacity, reference class ambiguity, and strategic interaction.
The judge is explicitly instructed not to equate proximity to 50\% with difficulty.
Presentation order is randomized within each pair to control for position bias.
After 15 rounds (745 total comparisons), we compute Elo ratings from the win--loss record using a $K$-factor of 32.

\paragraph{Linear mixed-effects model.}
We test whether Elo difficulty predicts belief sensitivity metrics beyond what $|p_0 - 0.5|$ already explains.
For each (model, question) pair, we compute per-question values of mean absolute shift, asymmetry index, and mean judge ratings (reasoning coherence and expressed uncertainty, each averaged across all probe ratings for that question).
We fit linear mixed-effects models of the form:
\begin{equation}
    y_{mq} = \beta_0 + \beta_1 \cdot \text{Elo}_q^{(z)} + \beta_2 \cdot |p_0 - 0.5|_q^{(z)} + u_m + \varepsilon_{mq}
\end{equation}
where $y_{mq}$ is the metric for model $m$ on question $q$, predictors are $z$-scored, $u_m \sim \mathcal{N}(0, \sigma_m^2)$ is a random intercept for model, and $\varepsilon_{mq}$ is residual error.
We assess the incremental contribution of Elo via likelihood ratio tests comparing the full model to a $|p_0 - 0.5|$-only baseline, and report marginal and conditional $R^2$ \cite{nakagawa2013general}.
As a validation check, we test whether Elo ratings correlate with inter-model disagreement (SD of initial probabilities across the seven models for the same question), controlling for $|p_0 - 0.5|$.

%\input{figures/elo_exemplars_table}

% ============================================================================

\subsection{Complexity Classification Prompt}
\label{app:complexity-prompt}

The following system prompt is used with GPT-5.4 (\texttt{openai/gpt-5.4}, temperature~0) to classify each ForecastBench question as HIGH or LOW causal complexity.

\begin{quote}
\small
\texttt{You are classifying forecasting questions by whether they have sufficient causal complexity to support a large (12-16 node) directed acyclic graph of interacting causal factors.}

\texttt{A question has HIGH causal complexity if ALL of the following are true:}
\begin{enumerate}
\small
\item \texttt{A domain expert would naturally identify 12+ distinct, non-redundant causal factors that INTERACT with each other (not just a linear chain)}
\item \texttt{The factors span 3+ domains (e.g., Political, Economic, Social, Technological, Military/security, Regulatory/legal, Environmental, Organizational)}
\item \texttt{The factors form a genuine causal NETWORK --- they influence each other, create indirect pathways, and have effects through mediating variables}
\item \texttt{The question is GROUNDED in real, identifiable actors, institutions, or measurable conditions with current observable evidence}
\item \texttt{The expert reasoning naturally involves understanding how factors interact to shape the outcome, not just checking a sequence of conditions}
\end{enumerate}

\texttt{A question is LOW if ANY of these apply:}

\texttt{Ungrounded/speculative: The outcome is speculative with no historical precedent; it depends on inventions or breakthroughs that do not yet exist; the causal factors would be entirely hypothetical.}

\texttt{Simple reasoning structure: It is primarily time-series extrapolation; the factors are mostly within a single domain; the factors form a linear chain; a reasonable analyst would rely on base rates, trends, or simple conditional reasoning.}

\texttt{Expert methodology shortcuts (use these as guides, not hard rules): Statistical/actuarial: base rates, trend extrapolation, time series. Tournament/competition: relative strength comparisons. Pipeline/stage-gate: linear progression through defined phases. Market pricing: outcomes aggregated into asset prices.}

\texttt{IMPORTANT: Some questions resolve as a single event but the CONDITIONS that determine that event emerge from a complex interacting network. If the context and drivers form a genuine multi-domain causal network --- even though the resolution is a discrete event --- classify as HIGH. The test is whether the DRIVERS interact in a network, not whether the outcome is binary.}
\end{quote}

% ============================================================================
\bibliography{references}
\bibliographystyle{acl_natbib}

\end{document}
